Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-08-20 Manuscript/manuscript.tex
radicals

The most well-known CI route is based on the excitation degree $e$.
Starting from a reference determinant, typically the Hartree-Fock (HF) determinant, one generates all connected determinants by exciting at most $e$ electrons.
The excitation degree thus defines the order of the approximate excitation-based CI (eCI) model:
CI with single excitations (CIS), CI with single and double excitations (CISD), CI with single, double, and triple excitations (CISDT), and so on.
The eCI route rather quickly captures dynamic (weak) correlation, though it struggles with the description of static (strong) correlation.

A different CI route is based on the seniority number $s$ (the number of unpaired electrons in a given determinant).
In seniority-based CI (sCI), \cite{Bytautas_2011,Allen_1962,Smith_1965,Veillard_1967} there is no reference determinant and one accounts for all determinants having seniority equal or less than $s$.
In contrast to eCI, sCI recovers static correlation more efficiently and works well to describe molecular dissociation, \cite{Bytautas_2015,Alcoba_2014,Alcoba_2014b}
at the expense of a poorer account of dynamic correlation and a higher computational cost.
For systems with an even number of electrons, the first approximate model is defined by $s=0$ (sCI0), usually referred to as doubly-occupied CI (DOCI),
which is followed by the higher-order models: sCI2, sCI4, and so on.
For odd numbers of electrons, the seniority route follows along odd numbers of $s$: sCI1, sCI3, and so on.

@@

The full set of PECs for the open-shell systems (\ce{OH}, \ce{CN}, \ce{H7}, and vinyl) are presented in the {\SupInf}.
From these PECs, we obtain the NPEs, distance errors, equilibrium bond lengths and harmonic vibrational frequencies, which are plotted as a function of $\Ndet$ in Fig.~\ref{fig:plot_all}.
Each point in the figure denotes one CI model.
For instance, the hCI route is represented by the dark green line, with the first point being HF, the second being hCI1, the third being hCI1.5, and so on.
The corresponding results for the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) are shown in Fig.~\ref{fig:plot_all_closed}.
When we first introduced hCI, \cite{Kossoski_2022} we surveyed the same molecules, though only for the bare CI models (without a perturbative correction).
These previous results are reproduced in Fig.~\ref{fig:plot_all_closed} together with the present results, which include such correction.
First we will discuss the results for the bare CI models (hCI, eCI and sCI) for the open-shell systems (represented by the dark tones in the figures),
leaving the discussion about the perturbative correction (light tones) for both open-shell and closed-shell systems to the next subsection.

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{OH}, \ce{CN}, vinyl, and \ce{H7},
as a function of the number of determinants ($\Ndet$), according to hCI (green), eCI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the standard EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

%%% FIG 4 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all_closed}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4}, and \ce{H8},
as a function of the number of determinants ($\Ndet$), according to hCI (green), eCI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the standard EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all_closed}
\end{figure*}
%%% %%% %%%

We found that hCI typically improves the convergence with respect to $\Ndet$ when compared to eCI, for the four properties surveyed here.
The present finding for open-shell systems therefore parallels the previously reported superiority of hCI for closed-shell systems. \cite{Kossoski_2022}

The smaller NPEs obtained with hCI are related to the better description of the PECs at dissociation.
This is due to a larger fraction of static correlation being recovered,
stemming from the classes of determinants appearing lower in the seniority-excitation map (see Fig.~\ref{fig:hCI}),
namely, determinants high in excitation degree $e$ and low in seniority $s$, absent in eCI of the same order.
The smaller NPEs attained with hCI are clear for \ce{OH} (involving a single bond breaking),
vinyl (double bond breaking) and \ce{CN} (triple bond breaking),
whereas for \ce{H7} (multiple bond breaking), hCI and eCI are comparable.
This dependence on the order of bond breaking had also been observed for closed-shell systems \cite{Kossoski_2022} (as can be seen in Fig.~\ref{fig:plot_all_closed}),
and would be expected, as describing dissociation becomes increasingly more challenging as the bond order increases.
Even though hCI manages to recover more static correlation than eCI,
the advantage of the former is more striking for single bond breaking, whereas multiple bond breaking (like \ce{H7} and \ce{H8}) inevitably requires higher order excited determinants.

@@

In great contrast to hCI and eCI, sCI models deliver a poor convergence of all observables.
The single exception is the NPE for \ce{OH} obtained from sCI0, in-between the NPEs obtained from hCI2 and CISDT, methods having a comparable computational cost in this particular case.
Going to larger basis sets or bigger systems, the computational burden of sCI models would increase considerably more than hCI or eCI models,
due to its formal exponential scaling.
As far as configuration interaction methods are concerned, we conclude that the sCI route is unattractive to open-shell systems,
in line with similar findings for closed-shell systems \cite{Kossoski_2022} (also see Fig.~\ref{fig:plot_all_closed}).

@@

We put forward the interesting perspective of developing hierarchy-based CC (hCC) methods (hCC1, hCC2, and so on),
along with derived EOM-CC formulations to target excited states (EOM-hCC1, EOM-hCC2, and so on).
EOM-CC with single and double excitations (EOM-CCSD) \cite{Rowe_1968,Monkhorst_1977,Koch_1990,Stanton_1993}
is very accurate for describing singly-excited states, \cite{Loos_2018,Loos_2020,Loos_2020a} albeit it slightly overestimates their excitation energies.
In contrast, CISD provides too large excitation energies, \cite{Koch_1990,Kossoski_2023}
even though both CISD and EOM-CCSD methods span the same excited determinants and rely on HF ground-state orbitals.
Here, we found that hCI2 is more accurate than CISD, despite their large absolute errors. In particular, the overestimated energies are less exaggerated in the former.
We ponder whether this improvement would be transferable to EOM-hCC2.
Along the same line, it would be interesting to develop and gauge the performance of cheaper methods like EOM-hCC1.
Importantly, it is not obvious what the computational scaling of such methods would be.
Alternatively, one can envision state-specific hCC methods to target excited states.

>>>>

The most well-known CI route is based on the excitation degree $e$.
Starting from a reference determinant, typically the Hartree-Fock (HF) determinant, one generates all connected determinants by exciting at most $e$ electrons.
The excitation degree thus defines the order of the approximate excitation-based CI (eCI) model:
CI with single excitations (CIS), CI with single and double excitations (CISD), CI with single, double, and triple excitations (CISDT), etc.
The eCI route rather quickly captures dynamic (weak) correlation, though it struggles with the description of static (strong) correlation.

A different CI route is based on the seniority number $s$ (the number of unpaired electrons in a given determinant).
In seniority-based CI (sCI), \cite{Bytautas_2011,Allen_1962,Smith_1965,Veillard_1967} there is no reference determinant and one accounts for all determinants having seniority equal or less than $s$.
In contrast to eCI, sCI recovers static correlation more efficiently and works well to describe molecular dissociation, \cite{Bytautas_2015,Alcoba_2014,Alcoba_2014b}
at the expense of a poorer account of dynamic correlation and a higher computational cost.
For systems with an even number of electrons, the first approximate model is defined by $s=0$ (sCI0), usually referred to as doubly-occupied CI (DOCI),
which is followed by the higher-order models: sCI2, sCI4, \ldots.
For odd numbers of electrons, the seniority route follows along odd numbers of $s$: sCI1, sCI3, and so on.

@@

The full set of PECs for the open-shell systems (\ce{OH}, \ce{CN}, \ce{H7}, and vinyl) are presented in the {\SupInf}.
From these PECs, we obtain the NPEs, distance errors, equilibrium bond lengths, and harmonic vibrational frequencies, which are plotted as functions of $\Ndet$ in Fig.~\ref{fig:plot_all}.
Each point in the figure denotes one CI model.
For instance, the hCI route is represented by the dark green line, with the first point being HF \titou{(or hCI0)}, the second being hCI1, the third being hCI1.5, etc.
The corresponding results for the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) are shown in Fig.~\ref{fig:plot_all_closed}.
When we first introduced hCI, \cite{Kossoski_2022} we surveyed the same molecules, though only for the bare CI models (without a perturbative correction).
These previous results are reproduced in Fig.~\ref{fig:plot_all_closed} together with the present results, which include such correction.
First, we discuss the results for the bare CI models (hCI, eCI, and sCI) for the open-shell systems (represented by the dark tones in the figures),
leaving the discussion about the perturbative correction (light tones) for both open-shell and closed-shell systems to the next subsection.

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{OH}, \ce{CN}, vinyl, and \ce{H7},
as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red), and sCI (blue) models,
with (light tones) and without (dark tones) the standard EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

%%% FIG 4 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all_closed}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4}, and \ce{H8}, as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the standard EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all_closed}
\end{figure*}
%%% %%% %%%

We find that hCI typically improves the convergence with respect to $\Ndet$ when compared to eCI, for the four properties surveyed here.
The present finding for open-shell systems, therefore, parallels the previously reported superiority of hCI for closed-shell systems. \cite{Kossoski_2022}

The smaller NPEs obtained with hCI are related to the better description of the PECs at dissociation.
This is due to a larger fraction of static correlation being recovered,
stemming from the classes of determinants appearing lower in the seniority-excitation map (see Fig.~\ref{fig:hCI}),
namely, determinants high in excitation degree $e$ and low in seniority $s$, absent in eCI of the same order.
The smaller NPEs attained with hCI are clear for \ce{OH} (involving a single bond breaking),
vinyl (double bond breaking), and \ce{CN} (triple bond breaking),
whereas for \ce{H7} (multiple bond breaking), hCI and eCI are comparable.
This dependence on the order of bond breaking had also been observed for closed-shell systems \cite{Kossoski_2022} (as can be seen in Fig.~\ref{fig:plot_all_closed}),
and would be expected, as describing dissociation becomes increasingly more challenging as the bond order increases.
Even though hCI manages to recover more static correlation than eCI,
the advantage of the former is more striking for single bond breaking, whereas multiple bond breaking (like \ce{H7} and \ce{H8}) inevitably requires higher-order excited determinants.

@@

In great contrast to hCI and eCI, sCI models deliver a poor convergence of all observables.
The single exception is the NPE for \ce{OH} obtained from sCI0, in between the NPEs obtained from hCI2 and CISDT, methods having a comparable computational cost in this particular case.
Going to larger basis sets or bigger systems, the computational burden of sCI models would increase considerably more than hCI or eCI models,
due to its formal exponential scaling.
As far as configuration interaction methods are concerned, we conclude that the sCI route is unattractive to open-shell systems,
in line with similar findings for closed-shell systems \cite{Kossoski_2022} (also see Fig.~\ref{fig:plot_all_closed}).
\PFL{Shall we comment on orbital optimization here?}

@@

We put forward the interesting perspective of developing hierarchy-based CC (hCC) methods (hCC1, hCC2, \ldots),
along with derived EOM-CC formulations to target excited states (EOM-hCC1, EOM-hCC2, \ldots).
EOM-CC with single and double excitations (EOM-CCSD) \cite{Rowe_1968,Monkhorst_1977,Koch_1990,Stanton_1993}
is very accurate for describing singly-excited states, \cite{Loos_2018,Loos_2020,Loos_2020a} albeit it slightly overestimates their excitation energies.
In contrast, CISD provides too large excitation energies, \cite{Koch_1990,Kossoski_2023}
even though both CISD and EOM-CCSD methods span the same excited determinants and rely on HF ground-state orbitals.
Here, we found that hCI2 is more accurate than CISD, despite their large absolute errors. In particular, the overestimated energies are less exaggerated in the former.
We ponder whether this improvement would be transferable to EOM-hCC2.
Along the same line, it would be interesting to develop and gauge the performance of cheaper methods like EOM-hCC1.
Importantly, it is not obvious what the computational scaling of such methods would be.
Alternatively, one can envision state-specific hCC methods to target excited states.
